---
title: "Assign isotypes"
author: Katie Evans
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

<br> 

* Group strains into new isotype groups based on pairwise concordance score.
* Match newly assigned groups to exisiting isotypes.
* Resolve problems:
    + whether any strains were assigned to > 1 new isotype groups
    + whether any existing isotypes matched to > 1 new isotype groups
    + whether any new group matched to > 1 old isotypes
* Plot new concordance values (red-blue) and old concordance values (red-yellow)

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo=FALSE)

# library(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyr)
library(magrittr)
library(DT)
library(plotly)
library(patchwork)
library(ComplexHeatmap)
library(circlize)
library(geosphere)
```

```{r}

##### UPDATE THESE PARAMETERS BEFORE RUNNING ######

# old gtcheck from previous release - to see if isotype assignments changed
gtcheck_old <- readr::read_tsv("20210121_gtcheck.tsv")

# new gt check from concordance-nf
gtcheck_new <- readr::read_tsv("concordance/gtcheck.tsv")

# sample coverage data - sample sheet for GATK
gatk_ss <- readr::read_tsv("/projects/b1059/analysis/wi-gatk/WI-20220125/sample_sheet_gatk_20210119.tsv") 

# existing isotype data - output from concordance (download google sheet for correct species)
WI_sheet <- readr::read_tsv("c_elegans_species_sheet_2022-02-01.tsv")

# concordance cutoff value
cutoff <- 0.9997

```

<br>

```{r, echo=F}
# define plot function
concordance_heatmap <- function(strains, version) {
    
    if (length(strains) > 2) {
        
        if (version == "new") {
            gtcheck_table <- gtcheck
            color_discordance = "royalblue1"
        }
        
        if (version == "old") {
            gtcheck_table <- gtcheck_old
            color_discordance = "yellow"
        }
        
        df1 <- gtcheck_table %>%
            dplyr::select(i, j, concordance) %>% 
            dplyr::filter(i %in% strains & j %in% strains) %>% 
            unique()
        
        df2 <- df1 %>% 
            dplyr::rename(i=j, j=i)
            
        df <- dplyr::bind_rows(df1, df2) %>% 
            unique() %>% 
            tidyr::spread(j, concordance) 
        
        if (nrow(df) > 2) {   
            
            if (nrow(df) <= 10) { font_size = 12}
            if (nrow(df) > 10 & nrow(df) < 15) { font_size = 10}
            if (nrow(df) >= 15 & nrow(df) < 30) { font_size = 8}
            if (nrow(df) >= 30 ) { font_size = 6}
            
            # make heatmap of concordance
            m = select(df, -i) %>% as.matrix()
            rownames(m) = df$i
            
            col_fun = circlize::colorRamp2(c(min(m, na.rm=T), max(m, na.rm=T)), c(color_discordance, "indianred1"))
            
            p  = ComplexHeatmap::Heatmap(m, col=col_fun, rect_gp = gpar(col = "white", lwd = 2), cell_fun = function(j, i, x, y, width, height, fill) {
                grid.text(sprintf("%.5f", m[i, j]), x, y, gp = gpar(fontsize = font_size)) 
            })
            
            print(p)
        } else { print("Not more than 2 strains in group in this dataset")}
        
    } else { print("Not more than 2 strains in group in this dataset")}
}

```



## Group strains based on concordance
```{r}
# Load previous gtcheck data for plotting and comparison
gtcheck_old <- gtcheck_old %>%
    dplyr::mutate(concordance = 1-(discordance/sites))

# Load coverage data
SM_coverage <- dplyr::select(gatk_ss, strain, coverage)

# Load existing isotype data
WI <- WI_sheet %>% 
    dplyr::select("strain", "isotype", "release", "isotype_ref_strain", "wgs_seq")

existing_isotype <- WI %>%
    dplyr::select(strain, isotype) %>%
    dplyr::filter(isotype != "NA") %>% unique()

# Calculate concordance and determine whether a strain pair is isotype
gtcheck <- gtcheck_new %>%
    dplyr::mutate(concordance = 1-(discordance/sites)) %>%
    dplyr::mutate(isotype = concordance > cutoff) 

# Create table of strain pairs that are isotype with each other
single_strains <- gtcheck %>%
    dplyr::select(i, j, isotype) %>%
    tidyr::gather(col, strain, -isotype) %>%
    dplyr::group_by(strain) %>%
    dplyr::summarise(single_strain = sum(isotype, na.rm = T)) %>% 
    dplyr::filter(single_strain==0)

single_strains <- as.data.frame(list(i = single_strains$strain,
                                     j = single_strains$strain,
                                     discordance = 0,
                                     concordance = 1,
                                     isotype = TRUE))

iso_gtcheck <- dplyr::bind_rows(gtcheck, single_strains) %>% 
    dplyr::filter(isotype == T)


# Generate isotype groups:
# For each strain, create a group that contains all strains that have "isotype==T" with this strain
stack_list <- function(x) {
    if (is.null(names(x)) == T) {
        names(x) <- as.character(1:length(x))
    }
    stack(x)
}

strain_list <- sort(unique(c(gtcheck$i, gtcheck$j)))

isotype_groups <- stack_list(unique(lapply(strain_list, function(x) {
    grouped_strains <- dplyr::filter(iso_gtcheck, (i == x | j == x)) %>%
        dplyr::select(i, j)
    sort(unique(unlist(grouped_strains)))
}))) %>%
    dplyr::rename(strain = values, group = ind) %>%
    dplyr::select(group, everything())


```

<br> 

## Check whether any strains were assigned to > 1 new isotype groups
```{r, fig.width=10, fig.height=8}
nrow(isotype_groups)
nrow(select(isotype_groups, strain) %>% unique())
nrow(select(isotype_groups, group) %>% unique())

multi = isotype_groups[duplicated(isotype_groups["strain"]) | duplicated(isotype_groups["strain"], fromLast = T), ]

datatable(multi, options = list(pageLength = 50), rownames=F)

# also look at all the strains these isotypes include
multi_iso = filter(isotype_groups, group %in% multi$group)

datatable(multi_iso, options = list(pageLength = 50), rownames=F)


if (nrow(multi_iso) > 0) {
    
    concordance_heatmap(multi_iso$strain, "new")
    concordance_heatmap(multi_iso$strain, "old")
    
}
```

<br>

## Match existing isotypes to each new group 
```{r, fig.width=10, fig.height=8}
# Match new group to exisiting isotypes by strain name
match_group_isotype <- dplyr::inner_join(isotype_groups, WI, by="strain") %>%  
    dplyr::left_join(SM_coverage) 

write.table(match_group_isotype, "isotype_group_match_0.9997.tsv", quote=F, sep="\t", col.names = T, row.names = F)

# Create new isotype group - exisiting isotype conversion table
match_group_isotype_unique <- dplyr::select(match_group_isotype, group, isotype) %>% 
    dplyr::filter(!is.na(isotype)) %>% 
    unique()

```


## Check whether any existing isotypes matched to > 1 new isotype groups
```{r}
multi2a = match_group_isotype_unique[ duplicated(match_group_isotype_unique$isotype) | duplicated(match_group_isotype_unique$isotype, fromLast = T), ]

datatable(multi2a, options = list(pageLength = 50), rownames=F)
```


```{r, fig.width=10, fig.height=8}
# plot
for (multi_group in unique(multi2a$group)) {
    
    print(paste0("Group ", multi_group))
    
    strains_in_multi2 = filter(isotype_groups, group==multi_group)
    
    concordance_heatmap(strains_in_multi2$strain, "new")
    concordance_heatmap(strains_in_multi2$strain, "old")
    
}
```


## Check whether any new group matched to > 1 old isotypes
```{r}
multi2 = match_group_isotype_unique[duplicated(match_group_isotype_unique["group"]) | duplicated(match_group_isotype_unique["group"], fromLast = T), ]

datatable(multi2, options = list(pageLength = 50), rownames=F)
```

<br>

```{r, fig.width=10, fig.height=8}

# plot
for (multi_group in unique(multi2$group)) {
    
    print(paste0("Group ", multi_group))
    
    strains_in_multi2 = filter(isotype_groups, group==multi_group)
    
    concordance_heatmap(strains_in_multi2$strain, "new")
    concordance_heatmap(strains_in_multi2$strain, "old")
    
}
```

<br>

### Plot ALL groups that match to > 1 old isotype
```{r, fig.width=10, fig.height=8}
strains_in_multi2 = filter(isotype_groups, group %in% multi2$group)

concordance_heatmap(strains_in_multi2$strain, "new")
concordance_heatmap(strains_in_multi2$strain, "old")

```


### Plot concordance in previous release of strains in these isotypes 
The previous plot are subsetted by strains, and this one is subsetted by existing isotypes. The result may be the same set of strains. 
```{r, fig.width=10, fig.height=8}
# get strains among these problematic isotypes in previous release
strains_in_multi2_isotypes = filter(WI, isotype %in% multi2$isotype)

concordance_heatmap(strains_in_multi2_isotypes$strain, "old")

```


## Plot other groups in question. 
Group that show introgression in within-group pairwise plots

```{r, fig.width=10, fig.height=8}
strains_in_multi2 = filter(isotype_groups, group=="")

concordance_heatmap(strains_in_multi2$strain, "new")
concordance_heatmap(strains_in_multi2$strain, "old")

```


<br>


## Zoom in new concordance
```{r}
# gtcheck %>%
#     # dplyr::filter(concordance < 0.9997) %>%
#     dplyr::mutate(strain_pair = paste0(i, "-", j)) %>%
#     ggplot(.) +
#     geom_histogram(aes(x=concordance), binwidth = 0.00001, fill="#0080FF") +
#     scale_x_continuous(limits = c(0.999, 1.0)) +
#     labs(x = "Concordance", y = "Number of Comparisons") +
#     geom_vline(aes(xintercept = cutoff), color = "red") +
#     theme(axis.title = ggplot2::element_text(size=14, face="bold", color="black", vjust=5))

test <- gtcheck %>%
    dplyr::mutate(bin = NA)

bins <- seq(0.999, 1.0, 0.00001)

# first bin
b <- bins[1]
test$bin[which(test$concordance > b)] <- b
for(i in 2:length(bins)) {
    b <- bins[i]
    # for each bin, which rows belong to that bin?
    test$bin[which(test$concordance > b)] <- b
}

test2 <- test %>%
    dplyr::mutate(strain_pair = glue::glue("{isotype}: {i}-{j} ({round(concordance, digits = 6)})")) %>%
    dplyr::group_by(bin) %>%
    dplyr::summarize(num = n(),
                     strain_text = paste(strain_pair, collapse = "\n")) %>%
    na.omit() %>%
    ggplot(.) +
    aes(x = bin, y = num) +
    geom_bar(stat = "identity", fill = "#0080FF") +
    labs(x = "Concordance", y = "Number of Comparisons") +
    geom_vline(aes(xintercept = cutoff), color = "red") +
    theme(axis.title = ggplot2::element_text(size=14, face="bold", color="black", vjust=5))

plotly::ggplotly(test2 + aes(text = glue::glue("{strain_text}")), tooltip = "text")


gtcheck %>% filter(isotype == TRUE) %>% 
    left_join(match_group_isotype, by=c("i"="strain")) %>% 
    left_join(match_group_isotype, by=c("j"="strain")) %>% 
    select(i, j, sites, discordance, concordance, group.x, isotype.y, isotype, coverage.x) %>% 
    mutate(coverage.x=as.integer(coverage.x)) %>%  
    unique() %>% 
    DT::datatable(colnames = c("Strain", "Strain", "Total variants", "Diff variants", "Concordance", "New isotype group", "i old isotype", "j old isotype", "Coverage"),
                  rownames = F,
                  options = list(pageLength = 50),
                  filter = 'top') # use scroller may change how filter look. be careful
```


## Zoom in previous concordance
```{r}
# ggplot(gtcheck_old) +
#     geom_histogram(aes(x=concordance), binwidth = 0.00001, fill="goldenrod") +
#     scale_x_continuous(limits = c(0.999, 1.0)) +
#     labs(x = "Concordance", y = "Number of Comparisons") +
#     geom_vline(aes(xintercept = 0.9997), color = "red") +
#     theme(axis.title = ggplot2::element_text(size=14, face="bold", color="black", vjust=5))

test <- gtcheck_old %>%
    dplyr::mutate(bin = NA)

bins <- seq(0.999, 1.0, 0.00001)

# first bin
b <- bins[1]
test$bin[which(test$concordance > b)] <- b
for(i in 2:length(bins)) {
    b <- bins[i]
    # for each bin, which rows belong to that bin?
    test$bin[which(test$concordance > b)] <- b
}

test2 <- test %>%
    dplyr::left_join(WI %>% dplyr::select(strain, isotype), by = c("i" = "strain")) %>%
    dplyr::left_join(WI %>% dplyr::select(strain, isotype), by = c("j" = "strain")) %>%
    dplyr::mutate(isotype = isotype.x == isotype.y) %>%
    dplyr::mutate(strain_pair = glue::glue("{isotype}: {i}-{j} ({round(concordance, digits = 6)})")) %>%
    dplyr::group_by(bin) %>%
    dplyr::summarize(num = n(),
                     strain_text = paste(strain_pair, collapse = "\n")) %>%
    na.omit() %>%
    ggplot(.) +
    geom_bar(aes(x = bin, y = num), stat = "identity", fill="goldenrod") +
    labs(x = "Concordance", y = "Number of Comparisons") +
    geom_vline(aes(xintercept = cutoff), color = "red") +
    theme(axis.title = ggplot2::element_text(size=14, face="bold", color="black", vjust=5))

plotly::ggplotly(test2 + aes(text = glue::glue("{strain_text}")), tooltip = "text")


```

<br> 

## Check location distance within isotype groups
```{r}
wi_loc <- WI_sheet %>%
    dplyr::select(strain, latitude, longitude)

p1 <- gtcheck %>% 
    dplyr::filter(concordance > 0.9992) %>% 
    dplyr::inner_join(wi_loc, by=c("i"="strain")) %>% 
    dplyr::inner_join(wi_loc, by=c("j"="strain")) %>% 
    dplyr::rowwise() %>% 
    dplyr::mutate(distance = geosphere::distm(c(longitude.x, latitude.x), c( longitude.y, latitude.y), fun = geosphere::distVincentyEllipsoid),
                  distance = distance / 1000) %>% 
    dplyr::mutate(label=paste0(i,"-",j)) %>% 
    ggplot2::ggplot() + 
        ggplot2::geom_point(aes(x=concordance, y=(distance+0.1), plotly_label=label, color = isotype), alpha=0.5, size=0.5) + 
        ggplot2::theme_light() + 
        ggplot2::ylab("Distance (km)") + 
        ggplot2::scale_color_manual(values = c("black", "blue")) +
        ggplot2::scale_y_continuous( labels = scales::comma, breaks = 10^seq(0 , 7, 1)) + 
        ggplot2::scale_x_continuous(breaks=seq(0.9992, 1, 0.0001))

ggplotly(p1, tooltip = "plotly_label")

gtcheck %>% 
    dplyr::filter(concordance > 0.9992) %>% 
    dplyr::inner_join(wi_loc, by=c("i"="strain")) %>% 
    dplyr::inner_join(wi_loc, by=c("j"="strain")) %>% 
    dplyr::rowwise() %>% 
    dplyr::mutate(distance = geosphere::distm(c(longitude.x, latitude.x), c( longitude.y, latitude.y), fun = geosphere::distVincentyEllipsoid),
                  distance = distance / 1000) %>% 
    dplyr::mutate(label=paste0(i,"-",j)) %>%
    dplyr::select(strain1 = i, strain2 = j, concordance, isotype, `strain1 latitude` = latitude.x, `strain1 longitude` = longitude.x, `strain2 latitude` = latitude.y, `strain2 longitude` = longitude.y, `distance (km)` = distance) %>%
    DT::datatable(rownames = F,
                  options = list(pageLength = 50),
                  filter = 'top') # use scroller may change how filter look. be careful

# plot for each isotype group, max distance value
test <- match_group_isotype %>% 
    dplyr::inner_join(wi_loc, by=c("strain"="strain")) %>% 
    dplyr::inner_join(wi_loc, by=c("isotype"="strain")) %>% 
    dplyr::rowwise() %>% 
    dplyr::mutate(distance = geosphere::distm(c(longitude.x, latitude.x), c( longitude.y, latitude.y), fun = geosphere::distVincentyEllipsoid),
                  distance = distance / 1000) %>%
    dplyr::group_by(isotype) %>%
    dplyr::summarise(max_dist = max(distance),
                     strains = paste(strain, collapse = "\n")) %>%
    dplyr::arrange(desc(max_dist))
test$isotype <- factor(test$isotype, levels = unique(test$isotype))

summary(test$max_dist)

p2 <- test %>%
    dplyr::filter(max_dist > 1) %>%
    ggplot2::ggplot() +
    aes(x = isotype, y = max_dist) +
    geom_bar(stat = "identity", fill = "navy") +
    theme_bw(10) +
    geom_hline(yintercept = 500, color = "red") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid = element_blank()) +
    labs(x = "Isotype", y = "Max strain distance (km)", title = "Isotype groups with at least one strain pair > 1 km distance")

plotly::ggplotly(p2 + aes(text = glue::glue("{strains}")), tooltip = "text")


```

```{r}

### might have to add some manual steps to remove and fix isotype groups

```


## Write output
<br>

### Assign isotypes to each strain
```{r}
# For all strains, reassign isotypes based on the group-isotype table generated above
final_isotype0 <- match_group_isotype %>% 
    dplyr::select(-isotype) %>%  # remove existing isotpyes 
    dplyr::left_join(match_group_isotype_unique, by="group") # re-assign isotype by group

 
# For strains that fall into groups with exisiting isotype reference strain, use that reference strain
final_isotype_not_NA <- dplyr::filter(final_isotype0, !is.na(isotype)) %>%
    dplyr::mutate(isotype_ref_strain = ifelse(is.na(isotype_ref_strain), FALSE, isotype_ref_strain)) 

# For strains without exisiting isotypes, use the strain with most coverage as istoype name and reference strain
final_isotype_NA <- dplyr::filter(final_isotype0, is.na(isotype)) %>%
    dplyr::select(-isotype)

NA_isotype <-  group_by(final_isotype_NA, group) %>%
    dplyr::mutate(max_cov = max(coverage)) %>%
    dplyr::filter(coverage==max_cov) %>% 
    dplyr::select(group, strain)

# Double check that each group only has 1 max coverage isotype
nrow(NA_isotype) == nrow(select(NA_isotype, group) %>% unique())

# Use the strain with max coverage as isotype name and ref strain
NA_isotype %<>% rename(isotype=strain)

final_isotype_NA %<>% left_join(NA_isotype, by="group") %>% dplyr::mutate(isotype_ref_strain = ifelse(strain==isotype, TRUE, FALSE))

# Combine existing and newly assigned isotypes
# remove 20201015 release date and change to 20210121
final_isotype <- dplyr::bind_rows(final_isotype_not_NA, final_isotype_NA) %>%
    dplyr::mutate(wgs_seq=TRUE) %>%
    dplyr::mutate(release=ifelse(is.na(release), release_date, release),
                  release = ifelse(release == "20201015", release_date, release)) %>%
    dplyr::select(strain, isotype, everything()) %>%
    dplyr::arrange(strain)

# Double check the total number of groups and isotypes are as expected
nrow(select(final_isotype, group) %>% unique) == nrow(select(final_isotype, isotype) %>% unique)
```

<br>


### Strain and isotype count

```{r}
# Existing
a <- select(final_isotype_not_NA, strain) %>% unique() %>% nrow()
b <- select(final_isotype_not_NA, group) %>% unique() %>% nrow()

print(paste0("Total of ", a, " strains assigned to ", b, " existing isotyes."))

# New
a <- select(final_isotype_NA, strain) %>% unique() %>% nrow()
b <- select(final_isotype_NA, group) %>% unique() %>% nrow()

print(paste0("Total of ", a, " strains assigned to ", b, " new isotyes."))
```

<br>

### Check that each isotype has 1 and only 1 ref strain
```{r}
nrow(final_isotype %>% 
    dplyr::group_by(isotype) %>% 
    dplyr::summarise(num_of_ref=sum(isotype_ref_strain)) %>%
    dplyr::filter(num_of_ref!=1)) == 0
```

<br>

### Check that all fields in table is filled
```{r}
is.na(final_isotype) %>% sum() == 0
```

<br>

### Final table
```{r}
datatable(final_isotype, rownames=F)
```

<br>

### Write out for merging into C.elegans strain sheet
```{r}
final_isotype %<>% dplyr::select( -group, -coverage)

write_tsv(final_isotype, "20210115_isotype_assignment.tsv")

```



### Write out list of isotype ref strains for subsetting
```{r}
final_isotype %>% 
    dplyr::filter(isotype_ref_strain) %>% 
    dplyr::select(strain) %>% 
    dplyr::inner_join(gatk_ss) %>%  
    readr::write_tsv("20210121_isotype_ref_strains.txt", col_names = F)
```



```{r}
sessionInfo()
```